D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-21 cuttingroom.tex
Drop "C++ Implementation" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

>>>>

\subsection{C++ Implementations}
\label{sec:C++ Implementations}

As noted earlier, this initial work focuses on C++ implementations based
on compilers and computing hardware (including CPUs and GPGPUs).

C++ implementations execute C++ programs, and use a range of techniques.
For example, many implementations take steps to optimize the code to
be executed.
This optimization presents a trade-off between the time spent optimizing
the code and the time spent actually executing it.
One type of optimization is the identification of nonsemantic dependencies
and the replacement of them with simpler computations, up to and including
replacement of arbitrarily complex passages of code with constants.\footnote{
	There are a great many other optimizations, but this paper
	necessarily focuses on semantic dependencies.}

This optimization thread may, roughly speaking, be divided into two
activities, analysis of the code and classification of dependencies.
Two sections on these topics are followed by sections on optimizations
and on consequences.

\subsubsection{Code Analysis}
\label{sec:Code Analysis}

As would be expected given the various optimization/execution tradeoffs,
different C++ implementations carry out different degrees of analysis.

First, at one extreme, ``simple implementations'' do absolutely no
analysis beyond that absolutely required to correctly execute the program.

Second, other implementations do local analysis.
This analysis might be on a per-code-path basis, within a single function,
within a single function after inlining, and so forth.

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe''), and loads from objects for which the value
returned by a given load was overwritten by some other thread
(``fre'').\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for more
	information on these memory-reference inter-process communications
	links.}

Fourth, one could also imagine implementations whose analysis is
completely unrestricted, up to and including the limiting case where
execution is oracular in nature.
Of course, if we are going to imagine omniscient analysis and oracles,
we can also rely on them to locate any OOTA cycles that might appear
in environments not constrained by the laws of physics, hardware
limitations, and C++ implementation concerns.

However, for more realistic implementations in the third category,
analysis relevant to OOTA cycles begins with loads from objects that were
last stored to by some other thread and ends with stores to objects that
might be loaded by other threads.

If additional inter-thread information is available, the corresponding
constraints may be applied.
For example, if an implementation proves that the values of a pair of
objects are always equal at any time that a given thread might access
them, that implementation could behave as if the that thread's code had
substituted a load from one for a load from the other.

Note that if all of the inter-thread relaxed accesses are volatile,
as they are in the C language, then all of the inter-thread traffic is
observed behavior.
In this case, even omniscient oracular implementations are required
to confine their analyses to within a single thread.

The next section focuses on the key analysis step for OOTA cycles,
namely dependency classification.

\subsubsection{Dependency Classification}
\label{sec:Dependency Classification}

Carrying out this type of optimization requires that the implementation
classify dependencies as maybe semantic or definitely not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as nonsemantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug, which is
most likely a bug even in the absence of OOTA cycles.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as nonsemantic, in which case
the execution is both correct and atemporal.

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Execution	& Implementation	& Result \\
\hline
sdep			& sdep			& Temporal \\
\cline{2-3}
			& $\neg$sdep		& Fix C++ bug! \\
\hline
$\neg$sde		& sdep			& Temporal \\
\cline{2-3}
			& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification}
\label{tab:Semantic-Dependency Classification}
\end{table}

Note well that in the absence of bugs in the implementation (which
again should be fixed), it is the implementation's classification of
the dependency that dictates whether or not the corresponding execution
is temporal.

\subsubsection{C++ Implementations: Optimization}
\label{sec:C++ Implementations: Optimization}

C++ implementations undertake a wide variety of optimizations, in
some cases taking advantage of nondeterminism and in others applying
the results of code analysis to produce faster code.
The changes can be profound, for example, with large computations being
replaced by simple constants.
Either way, the implementation is behaving as if the original program
had been written more optimally.

In such cases, one option is to carry out analysis of semantic dependencies
on the optimized code.
The seL4 project took this approach to the extreme of analyzing
the binary output~\cite{ThomasSewell2013L4binaryVerification}.

In C++ implementations using compilers, nondeterminism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other C++ implementations (for example, those using interpreters or JITs)
might need to directly model any nondeterminism.
This latter situation might motivate additional C++ projects to ban
nondeterminism in favor of portable code.

Another question is how to prove that a given optimization is valid.
This has been a hard question in the past, and is likely to remain so
both for sequential and concurrent programs.
The consequences of using an invalid optimization can be quite severe,
and the evidence thus far indicates that OOTA cycles are not a bit issue.
Nevertheless, a proof that the same value is always stored implies a
proof that there is no semantic dependency to that relaxed atomic store,
and thus no chance of that store participating in an OOTA cycle.

\subsubsection{C++ Implementations: Consequences}
\label{sec:C++ Implementations: Consequences}

Given a correct C++ implementation, a given store will either be
executed as if is was semantically dependent on a given prior load or
not.
In the former case on a real-world implementation, a nonnegative interval
of time will elapse between the load and the store, and in the latter
case, that load-store pair cannot possibly be a member of an OOTA cycle.

Of course, dependencies can be separately analyzed, and such analysis
might show that one of the dependencies that the implementation treated
as a semantic dependency was in fact nonsemantic.
And this is a belt-and-suspenders situation:
The corresponding load-store pair cannot be a member of an OOTA cycle,
and a nonnegative interval of time will elapse between the load and
the store.
